---
title: "ESM 244 Lab 1"
author: "Sophia Leiker"
date: "1/7/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

#Echo = FALSE --> removes code chunks and only shows output

library(tidyverse)
library(here)
library(sf)
library(tmap)

#to intsall packages --> install.packages('packagename') in console
```

cmd-shift-enter shortcut for running the current code chunk

### Reading in the data
Create a code chunk: command-option-i for creating a code chunk

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"))

#to see a summary of the dataset say summary(df) in the console
```

### Part 1: wragling and ggplot review

**Example 1:** Find counts of observations by `legal_status` and wrangle a bit.

```{r}
### Method 1: group_by() %>%  summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### Method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>%  #drop rows containing missing values in the legal_status column, if you leave this empty then it will drop any rows that contain a missing value
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% #moves columns around
  slice_max(tree_count, n = 5) %>% #this is taking the top 5 counts
  arrange(desc(tree_count)) #organize by descending order, 
 

#both of these methods do the same thing!
```

Make a graph of the top 5 from above

```{r}
ggplot(data = top_5_status, aes(x= fct_reorder(legal_status, tree_count), y = tree_count)) + #fct_reorder takes the legal status results and reorders them according to smallest to largest tree count
  geom_col(fill = 'darkgreen') + #fill changes the color of the columns
  labs(x = 'Legal Status', y = 'Tree Count') +
  coord_flip() + #flips so columns go from vertical alignment to horizontal alginment
  theme_minimal()
  
```


**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly
```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status %in% c('Permitted Site', 'Private') & caretaker %in% "MTA") #the row has to match all of the following arguments (if using a comma, for AND use &, for OR use |)

#remember if you use filter with  == c('a', 'b') it will look for a and then b, so in this case you have to use:
#filter(legal_status %in% c('Permitted Site', 'Private'))
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% #str_detect looks through the strings to see if any of the string is detected (aka  if the string has 'my name is Sophia' you can use str_detect to search only for Sophia)
  select(legal_status, date, lat = latitude, lon = longitude) #also renaming columns

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x=lon, y=lat)) +
  geom_point(color = 'darkgreen')
  
```

**Example 4:** use `tidyr::separate()`
Using separate to use the :: to separate the common name from the scientific name

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = " :: ")

#this is separating using the :: and renaming the two columns as listed above
```


** Example 5** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite("id_status", tree_id, legal_status, sep = "_COOL_")

#this is uniting in a new column "id_status"
# it is dropping the columns
```



